### 2024-06-06 journal/correctness.tex
some more edits from final proofreading

After proving some basic properties in Section \ref{sec::basicProperties},
we show in Section \ref{sec::propagating} that a double refresh at each node
suffices to propagate an operation to the root.
In Section \ref{sec::tracingCorrect} we show \op{GetEnqueue} and \op{IndexDequeue}
correctly navigate through the  tree.
Finally, we prove linearizability in Section \ref{sec::linearizability}.

@@

\begin{lemma} \label{lem::subblocksDistinct}
If $B$ and $B'$ are two \rev{different} blocks in nodes at the same depth in the ordering tree, their sets of subblocks are disjoint.
\end{lemma}
\begin{proof}
We prove the lemma by reverse induction on the depth.
If $B$ and $B'$ are in leaves, they have no subblocks, so the claim holds.
Assume the claim holds for nodes at depth $d+1$ and let $B$ and $B'$ be two blocks in nodes at depth $d$.
Consider the direct subblocks of $B$ and $B'$ defined by~(\ref{defsubblock}).
If $B$ and $B'$ are in different nodes at depth $d$, then their direct subblocks are disjoint.
\rev{Suppose $B$ and $B'$ are in the same node $v$.
By Lemma \ref{lem::headProgress}, the $\fld{end\sub{left}}$ values of blocks of $v$ are non-decreasing,
so the intervals of direct subblocks of different blocks in $v$ in $v$'s left child defined by (\ref{defsubblock}) are non-overlapping.
Similarly, the intervals of direct subblocks in $v$'s right child are non-overlapping.
In all cases, the direct subblocks of $B$ and $B'$} (at depth $d+1$) are disjoint, so the claim follows from the induction hypothesis.
\end{proof}

@@

%\Eric{I think this case was missing from the proof in the thesis}
Now suppose $R$'s call to \op{CreateBlock} returns \nl, causing $R$ to terminate at line \ref{addOP}.
Intuitively, since there are no operations in $\var{v}$'s children to promote, $op$ is already in $\var{v}$.
We formalize this intuition.
The value computed at line \ref{computeNumEnq} is
\begin{eqnarray*}
&&\var{num\sub{enq}}\\
&=&  \var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}} \\
&& + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}}\\
&&  - \var{v}.\fld{blocks}[h-1].\fld{sum\sub{enq}} \\
&=& \var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}}\\
&& + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}} \\
&&- \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft].\fld{sum\sub{enq}} \\
&& - \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright].\fld{sum\sub{enq}}.
\end{eqnarray*}
It follows from Invariant \ref{inv::sum} that $num\sub{enq}$ is the total  number of \enqueues\  in 
$\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft+1..B.\eleft]$ and
$\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright+1..B.\eright]$.
Similarly, $num\sub{deq}$ is the total number of \dequeues\ contained in these blocks.
Since $num\sub{enq}+num\sub{deq}=0$ at line \ref{testEmpty},
these blocks contain no operations.
By Corollary \ref{blockNotEmpty}, this means the ranges of blocks are empty, so that $\var{v}.\fld{blocks}[h-1].\eleft \geq \var{B}.\eleft \geq i$.
Hence, $B_\ell$ is already a subblock of some block in $\var{v}$, so $\var{v}$ contains $op$.
\end{proof}

@@

\begin{lemma}\label{lem::doubleRefresh}
Consider two consecutive terminating calls $R_1$, $R_2$ to \op{Refresh}($\var{v}$) by the same process.
All operations contained $\var{v}$'s children when $R_1$ begins
are contained in $\var{v}$ when $R_2$ terminates.
\end{lemma}
\begin{proof}
If either $R_1$ or $R_2$ performs a successful \op{CAS} at line \ref{cas} or terminates at line \ref{addOP}, the claim follows
from Lemma \ref{successfulRefresh}.
So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{cas}.
Let $h_1$ and $h_2$ be the values $R_1$ and $R_2$ read from $\var{v}.\head$ at line \ref{readHead}.
By Observation \ref{obs::headInc}, $h_2>h_1$.
By Lemma \ref{lem::headProgress}, $\var{v}.blocks[h_2]=\nl$ when $R_1$ executes line \ref{readHead}.
Since $R_2$ fails its \op{CAS} on $\var{v}.blocks[h_2]$, some other \op{Refresh} $R_3$ must have done
a successful \op{CAS} on $\var{v}.blocks[h_2]$ before $R_2$'s \op{CAS}.
$R_3$ must have executed line \ref{readHead} after $R_1$, since $R_3$ read the value $h_2$ from $\var{v}.\head$ and the value of $\var{v}.\head$ is non-decreasing, by Observation \ref{nonDecreasingHead}.
Thus, all operations contained in $\var{v}$'s children when $R_1$ begins
are also contained in $\var{v}$'s children when $R_3$ later executes line \ref{readHead}.
By Lemma \ref{successfulRefresh}, these operations are contained in $\var{v}$ when $R_3$ performs its successful \op{CAS},
which is before $R_2$'s failed \op{CAS}.
\end{proof}

@@

\begin{proof}
We prove the claim by induction on the depth of node $\var{v}$. The base case where $\var{v}$ is the root is trivial (see Line \ref{indexBaseCase}).
Assuming the claim holds for $\var{v}$'s parent, we prove it for $\var{v}$.
Let $B=\var{v}.\fld{blocks}[b]$ and $B'$ be the superblock of $B$.
\op{IndexDequeue}($\var{v}, b, i$) first computes the index $sup$ of $B'$ in $\var{v}.\fld{parent}$.
By Lemma \ref{superRelation}, this index is either $B.super$ or $B.super+1$.
The correct index is determined by testing on line \ref{supertest} whether $B$ is a subblock of $\var{v}.\fld{parent.blocks}[B.\fld{super}]+1$.

@@

Finally, \op{IndexDequeue} is called recursively on $\var{v}$'s parent.
Since $B$ has been propagated to the root, so has its superblock $B'$.
Thus, all preconditions of the recursive call are met.
By the induction hypothesis, the recursive call returns the location of the required \dequeue\ in the root.\end{proof}

@@

Assuming the claim holds for $\var{v}$'s children, we prove it for $\var{v}$.
Let $B$ be $\var{v}.\fld{blocks}[b]$.
By Equation (\ref{defSeqs}),
$E(B)$ is obtained by concatenating the \enqueue\ sequences of the direct subblocks
of $B$, which are listed in (\ref{defsubblock}).
By Invariant \ref{inv::sum}, $\var{sum\sub{left}}-\var{prev\sub{left}}$ is the number
of \enqueues\ in $E(B)$ that come from $B$'s subblocks in $\var{v}$'s left child.
Thus, $dir$ is set to the direction for the child of $\var{v}$ that contains the required \enqueue\ operation.
Moreover, when line \ref{endChooseDir} is reached, $i$ is the position of the required \enqueue\ within the portion $E'$ of $E(B)$ that comes from that child.
Thus,  line \ref{getChild} finds the index $b'$ of the subblock $B'$ containing the required \enqueue.
By Invariant \ref{inv::sum}, $\var{v}.\var{dir.blocks}[b'-1].\var{sum\sub{enq}} - \var{prev\sub{dir}}$ is the number of 
\enqueues\ in $E'$ before the \enqueues of block $B'$, so
the value $i'$ computed on line \ref{getChildIndex} is the position of the required \enqueue\ within $E(B')$.
Thus, the recursive call on line \ref{getRecurse} satisfies its precondition, and 
returns the required result, by the induction hypothesis.
\end{proof}

@@

\begin{proof}
We prove the claim by induction on $b$. 
The base case when ${b=0}$ is trivially true, since the queue is initially empty and 
$\var{root}.\fld{blocks}[0]$ contains an empty block whose \size\ field is $0$. 
Assuming the claim holds for $b-1$, we prove it for $b$.
The \fld{size} field of the block $B$ installed in $\var{root}.\fld{blocks}[b]$ is computed
at line \ref{computeLength} of a call to \op{CreateBlock}(\var{root, b}).
By the induction hypothesis, $root.\fld{blocks}[b-1].\size$ gives the size of the queue before the operations
of block $B$ are performed.
By Invariant \ref{inv::sum}, the values of \var{num\sub{enq}} and \var{num\sub{deq}}
are the number of \enqueues\ and \dequeues\ contained in $B$.
Hence, the size of the queue after the operations of $B$ are performed (with \enqueues\ before \dequeues\ as specified by~$L$)
is $\max(0, \var{root}.\fld{blocks}[b-1].\size + \var{num\sub{enq}} - \var{num\sub{deq}})$.
\end{proof}

>>>>

After proving some basic properties in Section \ref{sec::basicProperties},
we show in Section \ref{sec::propagating} that a double \op{Refresh} at each node
suffices to propagate an operation to the root.
In Section \ref{sec::tracingCorrect} we show \op{GetEnqueue} and \op{IndexDequeue}
correctly navigate through the  tree.
Finally, we prove linearizability in Section \ref{sec::linearizability}.

@@

\begin{lemma} \label{lem::subblocksDistinct}
If $B$ and $B'$ are two \rev{different} blocks in nodes at the same depth in the ordering tree, their sets of subblocks are disjoint.
\end{lemma}
\begin{proof}
We prove the lemma by reverse induction on the depth.
If $B$ and $B'$ are in leaves, they have no subblocks, so the claim holds.
Assume the claim holds for nodes at depth $d+1$ and let $B$ and $B'$ be two \rev{different} blocks in nodes at depth $d$.
Consider the direct subblocks of $B$ and $B'$ defined by~(\ref{defsubblock}).
If $B$ and $B'$ are in different nodes at depth $d$, then their direct subblocks are disjoint.
\rev{Suppose $B$ and $B'$ are in the same node $v$.
By Lemma \ref{lem::headProgress}, the $\fld{end\sub{left}}$ values of blocks of $v$ are non-decreasing,
so the intervals of direct subblocks of different blocks in $v$ in $v$'s left child defined by (\ref{defsubblock}) are non-overlapping.
Similarly, the intervals of direct subblocks in $v$'s right child are non-overlapping.
In all cases, the direct subblocks of $B$ and $B'$} (at depth $d+1$) are disjoint, so the claim follows from the induction hypothesis.
\end{proof}

@@

%\Eric{I think this case was missing from the proof in the thesis}
Now suppose $R$'s call to \op{CreateBlock} returns \nl, causing $R$ to terminate at line \ref{addOP}.
Intuitively, since there are no operations in $\var{v}$'s children to promote, $op$ is already in $\var{v}$.
We formalize this intuition.
The value computed at line \ref{computeNumEnq} is
\begin{eqnarray*}
&&\var{num\sub{enq}}\\
&=&  \var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}} \\
&& + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}}\\
&&  - \var{v}.\fld{blocks}[h-1].\fld{sum\sub{enq}} \\
&=& \var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}}\\
&& + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}} \\
&&- \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft].\fld{sum\sub{enq}} \\
&& - \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright].\fld{sum\sub{enq}}.
\end{eqnarray*}
It follows from Invariant \ref{inv::sum} that $num\sub{enq}$ is the total  number of \enqueues\  in 
$\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft+1..B.\eleft]$ and
$\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright+1..$ $B.\eright]$.
Similarly, $num\sub{deq}$ is the total number of \dequeues\ contained in these blocks.
Since $num\sub{enq}+num\sub{deq}=0$ at line \ref{testEmpty},
these blocks contain no operations.
By Corollary \ref{blockNotEmpty}, this means the ranges of blocks are empty, so that $\var{v}.\fld{blocks}[h-1].\eleft \geq \var{B}.\eleft \geq i$.
Hence, $B_\ell$ is already a subblock of some block in $\var{v}$, so $\var{v}$ contains $op$.
\end{proof}

@@

\begin{lemma}\label{lem::doubleRefresh}
Consider two consecutive terminating calls $R_1$, $R_2$ to \op{Refresh}($\var{v}$) by the same process.
All operations contained $\var{v}$'s children when $R_1$ begins
are contained in $\var{v}$ when $R_2$ terminates.
\end{lemma}
\begin{proof}
If either $R_1$ or $R_2$ performs a successful \op{CAS} at line \ref{cas} or terminates at line \ref{addOP}, the claim follows
from Lemma \ref{successfulRefresh}.
So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{cas}.
Let $h_1$ and $h_2$ be the values $R_1$ and $R_2$ read from $\var{v}.\head$ at line \ref{readHead}.
By Observation \ref{obs::headInc}, $h_2>h_1$.
By Lemma~\ref{lem::headProgress}, $\var{v}.blocks[h_2]=\nl$ when $R_1$ executes line~\ref{readHead}.
Since $R_2$ fails its \op{CAS} on $\var{v}.blocks[h_2]$, some other \op{Refresh} $R_3$ must have done
a successful \op{CAS} on $\var{v}.blocks[h_2]$ before $R_2$'s \op{CAS}.
$R_3$ must have executed line \ref{readHead} after $R_1$, since $R_3$ read the value $h_2$ from $\var{v}.\head$ and the value of $\var{v}.\head$ is non-decreasing, by Observation \ref{nonDecreasingHead}.
Thus, all operations contained in $\var{v}$'s children when $R_1$ begins
are also contained in $\var{v}$'s children when $R_3$ later executes line \ref{readHead}.
By Lemma \ref{successfulRefresh}, these operations are contained in $\var{v}$ when $R_3$ performs its successful \op{CAS},
which is before $R_2$'s failed \op{CAS}.
\end{proof}

@@

\begin{proof}
We prove the claim by induction on the depth of node $\var{v}$. The base case where $\var{v}$ is the root is trivial (see line \ref{indexBaseCase}).
Assuming the claim holds for $\var{v}$'s parent, we prove it for $\var{v}$.
Let $B=\var{v}.\fld{blocks}[b]$ and $B'$ be the superblock of $B$.
\op{IndexDequeue}($\var{v}, b, i$) first computes the index $sup$ of $B'$ in $\var{v}.\fld{parent}$.
By Lemma \ref{superRelation}, this index is either $B.super$ or $B.super+1$.
The correct index is determined by testing on line \ref{supertest} whether $B$ is a subblock of $\var{v}.\fld{parent.blocks}[B.\fld{super}]+1$.

@@

Finally, \op{IndexDequeue} is called recursively on $\var{v}$'s parent.
Since $B$ has been propagated to the root, its superblock $B'$ \rev{has also been propagated to the root}.
Thus, all preconditions of the recursive call are met.
By the induction hypothesis, the recursive call returns the location of the required \dequeue\ in the root.\end{proof}

@@

Assuming the claim holds for $\var{v}$'s children, we prove it for $\var{v}$.
Let $B$ be $\var{v}.\fld{blocks}[b]$.
By Equation (\ref{defSeqs}),
$E(B)$ is obtained by concatenating the \enqueue\ sequences of the direct subblocks
of $B$, which are listed in (\ref{defsubblock}).
By Invariant \ref{inv::sum}, $\var{sum\sub{left}}-\var{prev\sub{left}}$ is the number
of \enqueues\ in $E(B)$ that come from $B$'s subblocks in $\var{v}$'s left child.
Thus, $dir$ is set to the direction for the child of $\var{v}$ that contains the required \enqueue\ operation.
Moreover, when line \ref{endChooseDir} is reached, $i$ is the position of the required \enqueue\ within the portion $E'$ of $E(B)$ that comes from that child.
Thus,  line \ref{getChild} finds the index $b'$ of the subblock $B'$ containing the required \enqueue.
By Invariant \ref{inv::sum}, $\var{v}.\var{dir.blocks}[b'-1].\var{sum\sub{enq}} - \var{prev\sub{dir}}$ is the number of 
\enqueues\ in $E'$ before the \enqueues\ of block $B'$, so
the value $i'$ computed on line \ref{getChildIndex} is the position of the required \enqueue\ within $E(B')$.
Thus, the recursive call on line \ref{getRecurse} satisfies its precondition, and 
returns the required result, by the induction hypothesis.
\end{proof}

@@

\begin{proof}
We prove the claim by induction on $b$. 
The base case when ${b=0}$ is trivially true, since the queue is initially empty and 
$\var{root}.\fld{blocks}[0]$ contains an empty block whose \size\ field is $0$. 
Assuming the claim holds for $b-1$, we prove it for $b$.
The \fld{size} field of the block $B$ installed in $\var{root}.\fld{blocks}[b]$ is computed
at line \ref{computeLength} of a call to \op{CreateBlock}(\var{root, b}).
By the induction hypothesis, $root.\fld{blocks}[b-1].\size$ gives the size of the queue before the operations
of block $B$ are performed.
By Invariant \ref{inv::sum}, the values of \var{num\sub{enq}} and \var{num\sub{deq}}
\rev{computed on line \ref{computeNumEnq} and \ref{computeNumDeq}}
are the number of \enqueues\ and \dequeues\ contained in $B$, respectively.
Hence, the size of the queue after the operations of $B$ are performed (with \enqueues\ before \dequeues\ as specified by~$L$)
is $\max(0, \var{root}.\fld{blocks}[b-1].\size + \var{num\sub{enq}} - \var{num\sub{deq}})$, \rev{as described in Equation (\ref{size-eqn})}.
\end{proof}
